### 2015-03-30 swmpr_manu.Rnw
finished proofing

% Template for PLoS
% Version 3.1 February 2015
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
% the final text of your manuscript.
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig." instead of "Figure".
% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - tabs/spacing/line breaks within cells to alter layout or alignment
% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
% - colors, shading, or graphic objects
% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % %

@@

% packages I added
\usepackage{cleveref}
\usepackage[colorlinks=true,allcolors=Gray]{hyperref}
% \usepackage[hidelinks]{hyperref}
\usepackage[usenames,dvipsnames]{xcolor}

@@

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

@@

%knitr options
<<setup, cache = F, echo = F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, background = 'white', highlight = FALSE, prompt = TRUE)
options(replace.assign=TRUE,width=70,digits=1)
@

@@

Four core data elements are collected through the \ac{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.  The SWMPr package is developed for the continous abiotic monitoring network which includes a majority of the data collected by \ac{SWMP}.  Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).  Each reserve has no less than four water quality stations and one weather station at fixed locations.  Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.  All data are made accessible through the \ac{CDMO} web portal\footnote{\url{http://cdmo.baruch.sc.edu/}}, where multiple \ac{QAQC} measures are used to screen the information for accuracy and reliability.  The final data include all timestamped observations including relevant \ac{QAQC} flags with the appropriate qualifier.

@@

The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: retrieving, organizing, and analyzing.  Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class following standard S3 documentation methods \cite{Wickham14}.  Functions for organizing and analyzing the data provide methods for working with a \texttt{swmpr} object.  An additional group of `miscellaneous' functions are included as helpers for the main functions.  The following describes a general approach for using each category of functions.

@@

<<eval = F, cache = T>>=
# get nuts, wq, and met data as separate objects for the same reserve
data(apacpnut)
data(apacpwq)
data(apaebmet)
swmp1 <- apacpnut
swmp2 <- apacpwq
swmp3 <- apaebmet

# combine nuts and wq data by union
comb(swmp1, swmp2, method = 'union')

# combine nuts and wq data by intersect
comb(swmp1, swmp3, method = 'intersect')

# combine nuts, wq, and met data by nuts time series, two hour time step
comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
@

@@

The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).  The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages.  These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.  Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components. Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \texttt{ecometab} and \texttt{plot\_metab} functions.  The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.

The \texttt{aggregate} function aggregates parameter data for a \texttt{swmpr} object by set periods of observation.  This function is most useful for aggregating noisy data to evaluate trends on longer time scales, or to simply reduce the size of a dataset.  Data can be aggregated by years, quarters, months, weeks, days, or hours for a user-defined function, which defaults to the mean.  A \texttt{swmpr} object is returned for the aggregated data, although the \texttt{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.  Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects \href{http://cran.r-project.org/web/packages/data.table/index.html}{data.table} package.  This approach was used to facilitate plotting using predefined methods for Date and POSIX objects.  Additionally, the method of treating NA values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned (see the example below).  Finally, the default argument for \texttt{na.action} is set to \texttt{na.pass} for \texttt{swmpr} objects to preserve the time series of the input data.

@@

descrips <- c(
  'Aggregate \\texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  Aggregation function is user-supplied but defaults to mean.',
  'Aggregate metabolism data from a \\texttt{swmpr} object.  This is primarly used within \\texttt{plot\\_metab} but may be useful for simple summaries of raw daily data.',
  'Estimate ecosystem metabolism for a combined water quality and weatehr dataset using the open-water method.',
  'Decompose a \\texttt{swmpr} time series into trend, seasonal, and residual components.  This is a simple wrapper to \\texttt{decompose}.  Decomposition of monthly or daily trends is possible.',
  'Decompose a \\texttt{swmpr} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \\texttt{decompTs} in the wq package.  Only monthly decomposition is possible.',
  'Plot a histogram for a \\texttt{swmpr} object.',
  'Add lines to an existing \\texttt{swmpr} plot.',
  'Linearly interpolate missing data (\\texttt{NA} values) in a \\texttt{swmpr} object. The maximum gap size that is interpolated is defined as a maximum number of records with missing data.', 
  'Plot a univariate  time series for a \\texttt{swmpr} object.  The parameter name must be specified.',
  'Plot ecosystem metabolism estimates after running \\texttt{ecometab} on a \\texttt{swmpr} object.',  
  'Create summary plots of seasonal/annual trends and anomalies for a water quality or weather parameter.',
  'Smooth \\texttt{swmpr} objects with a moving window average.  Window size and sides can be specified, passed to \\texttt{filter}.'
  )

@@

<<eval = F, cache = T>>=
# combine, qaqc, remove empty columns
dat <- comb(swmp1, swmp2, method = 'union')
dat <- qaqc(dat)
swmpr_in <- subset(dat, rem_cols = T)

# get mean DO by quarters
aggregate(swmpr_in, 'quarters', params = c('do_mgl'))

# get mean DO by quarters, remove NA when calculating means
fun_in <- function(x) mean(x, na.rm = T)
aggregate(swmpr_in, FUN = fun_in, 'quarters', params = c('do_mgl'))
@

Time series can be smoothed to better characterize a signal independent of noise (\Cref{fig:smooth_ex}).  Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.  The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size.  This method is a simple wrapper to \texttt{filter}.  The \texttt{window} argument specifies the number of observations included in the moving average.  The \texttt{sides} argument specifies how the average is calculated for each observation (see the documentation for \texttt{filter}).  A value of 1 will filter observations within the window that are previous to the current observation, whereas a value of 2 will filter all observations within the window centered at zero lag from the current observation. As before, the \texttt{params} argument specifies which parameters to smooth.  See \Cref{fig:smooth_ex} for the output from the code.

<<smooth_ex, eval = T, fig.height = 3, cache = T, fig.cap = "Raw and smoothed dissolved oxygen data for a two-week period after using the \\texttt{smoother} function.">>=
# import data
data(apadbwq)
swmp1 <- apadbwq

# qaqc and subset imported data
dat <- qaqc(swmp1)
dat <- subset(dat, subset = c('2012-07-09 00:00', '2012-07-24 00:00'))

# filter
test <- smoother(dat, window = 50, params = 'do_mgl')

# plot to see the difference
plot(do_mgl ~ datetimestamp, data = dat, type = 'l')
lines(test, select = 'do_mgl', col = 'red', lwd = 2)
@

A common issue with any statistical analysis is the treatment of missing values.  Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.  In either case, an analyst should have a strong rationale for the chosen method.  A common approach used to handle missing data in time series analysis is linear interpolation.  A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.  A challenge with linear interpolation is an appropriate gap size for fitting missing observations.  The ability of the interpolated data to approximate actual trends is a function of the gap size.  Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are more likely to resemble actual patterns.  An appropriate gap size limit depends on the unique characteristics of specific datasets or parameters.  The \texttt{na.approx} function can be used to interpolate gaps in a \texttt{swmpr} object.  A required argument for the function is \texttt{maxgap} which defines the maximum gap size for interpolation.  See \Cref{fig:interp_ex} for the output from the following code.

<<interp_ex, eval = T, fig.height = 6, cache = T, fig.scap = "Examples illustrating use of the \\texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.", fig.cap = "Examples illustrating use of the \\texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.">>=
# get data
data(apadbwq)
swmp1 <- apadbwq

# qaqc and subset imported data
dat <- qaqc(swmp1)
dat <- subset(dat, subset = c('2013-01-22 00:00', '2013-01-26 00:00'))

# interpolate, maxgap of 10 records
test <- na.approx(dat, params = 'do_mgl', maxgap = 10)

# interpolate maxgap of 30 records
test2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)

# plot for comparison
par(mfrow = c(3, 1))
plot(do_mgl ~ datetimestamp, dat, main = 'Raw', type = 'l')
plot(do_mgl ~ datetimestamp, test, col = 'red', 
  main = 'Interpolation - maximum gap of 10 records', type = 'l')
lines(dat, select = 'do_mgl')
plot(do_mgl ~ datetimestamp, test2, col = 'red', 
  main = 'Interpolation - maximum gap of 30 records', type = 'l')
lines(dat, select = 'do_mgl')
@

The \texttt{decomp} function is a simple wrapper to \texttt{decompose} that separates a time series into additive or multiplicative components describing a trend, cyclical variation (e.g., daily or seasonal), and the remainder.  The additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), whereas a multiplicative decomposition accounts for non-stationarity.  By default, a moving average with a symmetric window is used to filter the seasonal component.  Alternatively, a vector of filter coefficients in reverse time order can be supplied (see the help documentation for \texttt{decompose}).

The \texttt{decompose} function requires a \texttt{ts} object with a specified frequency as input.  The \texttt{decomp} function converts the input \texttt{swmpr} vector to a \texttt{ts} object prior to \texttt{decompose}.  This requires an explicit input defining the frequency of the parameter in the time series.  For example, the frequency of a parameter with diurnal periodicity would be 96 if the time step is 15 minutes (4 * 24).  The frequency of a parameter with seasonal periodicity would be 35040 (4 * 24 * 365).  For simplicity, character strings of \texttt{`daily'} or \texttt{`seasonal'} can be supplied in place of numeric values.  A starting value of the time series must be supplied in the latter case.  Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.  Note that the \texttt{decompose} function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.  \Cref{fig:decomp_ex1} is an example of the \texttt{decomp} function.

@@

An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package.  Theory describing this method is provided by Cloern and Jassby \cite{Cloern10}.  The function is similar to \texttt{decomp.swmpr} with a few key differences.  The \texttt{decomp.swmpr} function decomposes the time series into a trend, seasonal, and random component, whereas the current function decomposes into the grandmean, annual, seasonal, and events components.  For both functions, the random or events components, respectively, can be considered anomalies that do not follow the trends in the remaining categories.  The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.  This approach works best for nutrient data that are typically obtained on a monthly cycle.  The function will also work with continuous water quality or weather data but note that the data must first aggregated on the monthly scale before decomposition.  Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.  Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.  \Cref{fig:decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.

@@

# decomposition of chl, ggplot
decomp_cj(dat, param = 'chla_n')
@

Several graphics showing seasonal and annual trends for a given SWMP parameter can be obtained using the \texttt{plot\_summary} function.  The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.  Anomalies are defined as the difference between the monthly or annual average from the grand mean for the parameter.  Monthly anomalies are in relation to the grand mean for the same month across all years.  All data are aggregated for quicker plotting.  Nutrient data are based on monthly averages, whereas weather and water quality data are based on daily averages.  Cumulative precipitation data are based on the daily maximum. The function returns a graphics object (Grob) of multiple ggplot objects.  An interactive Shiny application \cite{shiny15} that uses this function is available (see the \nameref{supp_info}).

@@

Estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.  The open-water method \cite{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model (see Ro and Hunt \cite{Ro06}, Thebault et al. \cite{Thebault08}). The diffusion-corrected \ac{DO} flux estimates are averaged separately over each day and night of the time series. The nighttime average \ac{DO} flux is used to estimate respiration rates, while the daytime \ac{DO} flux is used to estimate net primary production. To generate daily integrated rates, respiration rates are assumed constant such that hourly night time \ac{DO} flux rates are multiplied by 24. Similarly, the daytime \ac{DO} flux rates are multiplied by the number of daylight hours, which varies with location and time of year, to yield net daytime primary production. Respiration rates are subtracted from daily net production estimates to yield gross production rates. The metabolic day is considered the 24 hour period between sunsets on two adjacent calendar days

The \texttt{ecometab} function is used to implement the open-water method with a combined water quality and weather dataset \cite{Caffrey14}.  Several assumptions must be met for a valid interpretation of the results.  In general, the dissolved oxygen time series is assumed to represent the same water mass over time.  Tidal advection may have a significant influence on the time series, which can contribute to a significant amount of noise in metabolic estimates.  The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.  Areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).  Water column depth is calculated as the mean value of the depth variable across the time series in the \texttt{swmpr} object. Depth values are floored at one meter for very shallow stations and 0.5 meters is also added to reflect the practice of placing sensors slightly off of the bottom.  Additionally, the air-sea gas exchange model is calibrated with wind data either collected at, or adjusted to, wind speed at 10 m above the surface.  The metadata should be consulted for exact height. Other assumptions may apply and relevant resources should be consulted to ensure appropriate application of the open-water method (see \cite{Kemp12,Needoba12}).

The following is an example that shows use of the function from a combined water quality and weather data set.  The results can be plotted using \texttt{plot\_metab} (\Cref{fig:metab_ex}).

<<metab_ex, eval = TRUE, cache = TRUE, fig.height = 4, fig.width = 8, warning = FALSE, fig.cap = "Monthly aggregations of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data at Apalachicola Bay, Florida.">>=
## import water quality and weather data
data(apadbwq)
data(apaebmet)

@@

latex(
  to_tab[, 'Description', drop = F],
  file = '',
  caption = "Miscellaneous functions available from the SWMPr package.  Most are used within the main functions above but may be useful for more customized evaluationg of \\ac{SWMP} data.  Full documentation for each function is in the help file (e.g., execute \\texttt{?calckl} at the command line).",
  rowlabel = 'Function',
  colheads = 'Description',
  rowname = to_tab$Functions,
  caption.loc = 'top',
  col.just=c("p{3.5in}"), 
  label = 'tab:misc', 
  table.env = FALSE
  )

>>>>

% Template for PLoS
% Version 3.1 February 2015
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig." instead of "Figure".
% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - tabs/spacing/line breaks within cells to alter layout or alignment
% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
% - colors, shading, or graphic objects
% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
%
% % % % % % % % % % % % % % % % % % % % % % %

@@

% packages I added
\usepackage{cleveref}
\usepackage[colorlinks=true,allcolors=Gray]{hyperref}
\usepackage[usenames,dvipsnames]{xcolor}

@@

@@

%knitr options
<<setup, cache = F, echo = F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, background = 'white', highlight = FALSE, prompt = TRUE, size = 'small')
options(replace.assign=TRUE,width=70,digits=1)
@

@@

Four core data elements are collected through the \ac{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.  The SWMPr package is developed for the continuous abiotic monitoring network which includes a majority of the data collected by \ac{SWMP}.  Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).  Each reserve has no less than four water quality stations and one weather station at fixed locations.  Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.  All data are made accessible through the \ac{CDMO} web portal\footnote{\url{http://cdmo.baruch.sc.edu/}}, where multiple \ac{QAQC} measures are used to screen the information for accuracy and reliability.  The final data include all timestamped observations including relevant \ac{QAQC} flags with the appropriate qualifier.

@@

The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: retrieving, organizing, and analyzing.  Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class following standard S3 documentation methods \cite{Wickham14}.  Functions for organizing and analyzing the data provide methods for working with a \texttt{swmpr} object.  An additional group of `miscellaneous' functions are included as helpers for the main functions.  The following describes a general approach for working with \ac{SWMP} data using each category of functions.

@@

<<eval = F, cache = T>>=
# get nut, wq, and met data as separate objects
data(apacpnut)
data(apacpwq)
data(apaebmet)
swmp1 <- apacpnut
swmp2 <- apacpwq
swmp3 <- apaebmet

# combine nut and wq data by union
comb(swmp1, swmp2, method = 'union')

# combine nut and wq data by intersect
comb(swmp1, swmp3, method = 'intersect')

# combine nut, wq, and met data by nut time series, two hour time step
comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
@

@@

The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).  The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages.  These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.  Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components. Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \texttt{ecometab} and \texttt{plot\_metab} functions.  The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.

The \texttt{aggregate} function aggregates parameter data for a \texttt{swmpr} object by set units of time.  This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.  Data can be aggregated by years, quarters, months, weeks, days, or hours by a predefined function, which defaults to the mean.  A \texttt{swmpr} object is returned for the aggregated data, although the \texttt{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.  Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects created in the data.table package \cite{Dowle14}.  Additionally, the method of treating \texttt{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data.

@@

descrips <- c(
  'Aggregate \\texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.',
  'Aggregate metabolism data from a \\texttt{swmpr} object.  This is primarly used within \\texttt{plot\\_metab} but may be useful for simple summaries of raw metabolism data.',
  'Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.',
  'Decompose a \\texttt{swmpr} time series into trend, seasonal, and residual components.  This is a simple wrapper to \\texttt{decompose} \\cite{Kendall83}.  Decomposition of monthly or daily trends is possible.',
  'Decompose a \\texttt{swmpr} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \\texttt{decompTs} in the wq package \\cite{Jassby14}.  Only monthly decomposition is possible.',
  'Plot a histogram for a \\texttt{swmpr} object.',
  'Add lines to an existing plot created with \\texttt{plot.swmpr}.',
  'Linearly interpolate missing data (\\texttt{NA} values) in a \\texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.', 
  'Plot a univariate  time series for a \\texttt{swmpr} object.  The parameter name must be specified.',
  'Plot ecosystem metabolism estimates after running \\texttt{ecometab} on a \\texttt{swmpr} object.',  
  'Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.',
  'Smooth \\texttt{swmpr} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \\texttt{filter}.'
  )

@@

<<eval = F, cache = T>>=
# get data
data(apacpwq)
dat <- qaqc(apacpwq)

# get mean DO by months
aggregate(dat, 'months', params = c('do_mgl'))
@

Time series can be smoothed to better characterize a signal independent of noise (\Cref{fig:smooth_ex}).  Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.  The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size with arguments passed to the \texttt{filter} function from the base installation.  The \texttt{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.  The \texttt{sides} argument specifies how the average is calculated for each observation.  Setting \texttt{sides = 1} will filter observations within the window that are previous to the current observation, whereas \texttt{sides = 2} will filter observations within the window centered at zero lag from the current observation. As before, the \texttt{params} argument specifies which parameters to smooth.

<<smooth_ex, eval = T, fig.height = 3, cache = T, fig.cap = "Raw and smoothed dissolved oxygen data for a two-week period after using the \\texttt{smoother} function.">>=
# import data, qaqc and subset
data(apadbwq)
dat <- qaqc(apadbwq)
dat <- subset(dat, select = 'do_mgl', 
  subset = c('2012-07-09 00:00', '2012-07-24 00:00')
  )

# smooth
dat_smooth <- smoother(dat, window = 50, params = 'do_mgl')

# plot raw and smoothed
plot(dat, type = 'l')
lines(dat_smooth, select = 'do_mgl', col = 'red', lwd = 2)
@

A common issue with any statistical analysis is the treatment of missing values.  Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.  In either case, an analyst should have a strong rationale for the chosen method.  A common approach implemented in the SWMPr package is linear interpolation using the \texttt{na.approx} function (\Cref{fig:interp_ex}).  A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.  However, the ability of the interpolated data to approximate actual trends is related to the maximum gap size between observations with missing data.  Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are often more accurate.  An upper limit on the maximum gap size to interpolate trends depends on the characteristics of the dataset such that a trial and error approach is appropriate for most applications.  The \texttt{maxgap} argument passed to \texttt{na.approx} defines the maximum gap size for interpolation and the following illustrates use of different maximum values to fill missing data.

<<interp_ex, eval = T, fig.height = 6, cache = T, fig.scap = "Examples illustrating use of the \\texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.", fig.cap = "Examples illustrating use of the \\texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.">>=
# get data, qaqc and subset
data(apadbwq)
dat <- qaqc(apadbwq)
dat <- subset(dat, select = 'do_mgl', 
  subset = c('2013-01-22 00:00', '2013-01-26 00:00'))

# interpolate, maxgap of 10 records
fill1 <- na.approx(dat, params = 'do_mgl', maxgap = 10)

# interpolate maxgap of 30 records
fill2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)

# plot for comparison
par(mfrow = c(3, 1))
plot(dat, main = 'Raw')
plot(fill1, col = 'red', main = 'Interpolation - maximum gap of 10 records')
lines(dat)
plot(fill2, col = 'red', main = 'Interpolation - maximum gap of 30 records')
lines(dat)
@

The disaggregation of time series into additive of multiplicative components that can be attributed to separate sources of variance is another common application for trend analysis.  The \texttt{decomp} function is a simple wrapper to \texttt{decompose} \cite{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (\Cref{fig:decomp_ex1}).  An additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), otherwise a multiplicative decomposition can be used.  The \texttt{frequency} argument describes the periodicity of the cyclical parameter in units of the native time step.  For example, the \texttt{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (4 * 24).  The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (4 * 24 * 365).  For simplicity, character strings of \texttt{`daily'} or \texttt{`annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.  A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation (e.g., 1 if the the first observation is at sunrise for a diurnal cycle).\footnote{The value is passed to the \texttt{ts} function, \url{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}}  Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.

@@

An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package \cite{Cloern10,Jassby14}.  The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.  This approach works best for nutrient data that are typically obtained on a monthly cycle.  The function will also work with continuous water quality or weather data but note that the data must first be aggregated on the monthly scale before decomposition.  Additionally, the time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \texttt{decomp} function described above.  For both, the random or events components, respectively, can be considered anomalies that do not follow the trends in the remaining categories.  Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.  Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.  \Cref{fig:decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.

@@

# decomposition of chl
decomp_cj(dat, param = 'chla_n')
@

Detailed exploratory graphics are also useful for evaluating general trends in observed data.  Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \texttt{plot\_summary} function (\Cref{fig:summary_ex}).  The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.  Anomalies are defined as the difference between the monthly or annual averages from the grand mean for the parameter.  An interactive web application \cite{Chang15} that uses this function is available for viewing results of any parameter at all \ac{SWMP} sites (see the \nameref{supp_info}).

@@

Finally, estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.  The open-water method \cite{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \cite{Ro06,Thebault08}. The diffusion-corrected \ac{DO} flux estimates are averaged separately over each day and night of the time series. The nighttime average \ac{DO} flux is used to estimate respiration rates, while the daytime \ac{DO} flux is used to estimate net primary production. To generate daily integrated rates, respiration rates are assumed constant such that hourly night time \ac{DO} flux rates are multiplied by 24. Similarly, the daytime \ac{DO} flux rates are multiplied by the number of daylight hours, which varies with location and time of year, to yield net daytime primary production. Respiration rates are subtracted from daily net production estimates to yield gross production rates. The metabolic day is considered the 24 hour period between sunsets on two adjacent calendar days

The \texttt{ecometab} function is used to implement the open-water method with a combined water quality and weather dataset \cite{Caffrey14}.  Several assumptions must be met for a valid interpretation of the results.  First, the \ac{DO} time series is assumed to be a sample of the same water mass over time.  Tidal advection may have a significant influence on the time series, which can contribute to a significant amount of noise in metabolic estimates.  The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.  Second, areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).  Water column depth is calculated as the mean value of the depth variable across the time series in the \texttt{swmpr} object. Depth values are floored at one meter for very shallow stations and 0.5 meters is also added to reflect the practice of placing sensors slightly off of the bottom.  Third, the air-sea gas exchange model is calibrated with wind data either collected at, or adjusted to, wind speed at 10 m above the surface.  The metadata should be consulted for exact height. Other assumptions may apply and relevant resources should be consulted to ensure appropriate application of the open-water method (see \cite{Kemp12,Needoba12}).

The following is an example that shows use of the function from a combined water quality and weather data set.  Monthly aggregations of the raw, daily estimates are plotted using \texttt{plot\_metab} (\Cref{fig:metab_ex}).

<<metab_ex, eval = TRUE, cache = TRUE, fig.height = 4, fig.width = 8, warning = FALSE, fig.cap = "Monthly means (95\\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida.">>=
## import water quality and weather data
data(apadbwq)
data(apaebmet)

@@

latex(
  to_tab[, 'Description', drop = F],
  file = '',
  caption = "Miscellaneous functions available from the SWMPr package.  Most are used within the main functions above but may be useful for customized evaluations of \\ac{SWMP} data.  Full documentation for each function is in the help file (e.g., execute \\texttt{?calckl} at the command line).",
  rowlabel = 'Function',
  colheads = 'Description',
  rowname = to_tab$Functions,
  caption.loc = 'top',
  col.just=c("p{3.5in}"), 
  label = 'tab:misc', 
  table.env = FALSE
  )
